{\bf Code Generation Recovery Log Replay for In-Memory Database Management Systems}

### 2021-08-04 paper.tex

\hypersetup{colorlinks,linkcolor={blue},citecolor={blue},urlcolor={red}}

@@

\begin{abstract}
Code generation is an optimization technique for improving query execution throughput by compiling query plans into native code. This technique, however, leads to design challenges for the recovery system of a database management system. The log replay process will be disjoint from the built-in execution engine that has adapted to operate efficiently on compiled code. This leads to the implementation of a separate execution engine to deal with the execution of log records, which can be a huge waste of engineering efforts. To resolve this design conflict between code generation and database recovery, we present a unified approach to support both query execution and log replay in a code-generation-based DBMS. We ask the recovery system to convert log records into compiled code that will be easily accepted by the execution engine. Our results show that while our approach incurs a higher performance overhead compared to using a separate execution engine, it requires much less engineering effort and is potentially superior in aspects such as index updates and constraint checking.
\end{abstract}

@@

Moreover, the DBMS recovery system is isolated from the query compilation execution engine (i.e., a component of a DBMS that is responsible for execution). Most of the query compilation DBMSs we have surveyed~\cite{noisepage, neumann11, paroski16, craig14} have no conduits between their recovery systems and execution pipelines. The recovery system implements its own functionalities to update the database, but some of these functionalities already exist in the execution engine. For instance, we mentioned that in-memory DBMSs rebuild the indexes during log replay. When a redo process install changes on a table, it also retrieves the indexes on that table and update them (index update). However, the execution engine has already implemented this index update functionality. As a result, the DBMS contains two implementations that provide identical functionalities. Similar reasoning applies to other functionalities (e.g., constraint checking, trigger events) that may be used during log replay. The number of those functionalities in the recovery system will eventually amount to the scale of a separate execution engine that is built specifically for log replay.

@@

\section{Transactions}
Transactions are the atomic operations of any DBMSs~\cite{theo83}. They appear throughout a DBMS, including the recovery system and the execution engine responsible for code generation. A transaction is a sequence of interactions with the database. These interactions can be retrieving data, inserting data, removing data, etc.

@@

There are two solutions to this problem. One solution is to re-implement functions to install physical changes on data tables within the recovery system. However, this approach completely isolates the recovery system from the execution engine. This means that if the recovery system wishes to use a functionality within the execution engine, it has to rewrite that same functionality within the recovery system (\cref{fig:problem}).

@@

If the DBMS can convert log records into physical plans, the recovery system can rely on code generation for recovery operation. This code generation recovery approach brings around several benefits. Firstly, it significantly reduces engineering overhead. Compiling contents of the log records into machine code allows the recovery system to seamlessly integrate with the built-in execution engine. We have explained in Section 2.1 that the recovery system needs to perform extra operations (e.g. index updates, constraint checking in \cref{fig:problem}) besides updates to the data tables. These functionalities exist within the execution engine. If we allow the recovery system to reuse existing functionalities by communicating with the execution engine through native code, then the recovery system no longer needs to implement its own functionalities. Secondly, utilizing the code generation execution engine provides more efficient functionalities. A function implemented by the recovery system will not be as efficient as one that exists in the execution engine.

These potential benefits motivate us to design a code generation recovery system based on conversion from log records to physical plans. In the next chapter, we will present the implementation of our approachhbmmn .

@@

We now describe how to extend the recovery system and query compilation pipeline to support log replay with code generation. Our implementation follows along the third option shown in \cref{fig:motivation}. We first enable the recovery system to convert log records into compiled native code. To achieve this, the DBMS needs to interpret the log records into physical plans and prepare them for compilation. We then need to make minor modifications to the execution engine, so it will be able to correctly execute the native code generated by the recovery system. We further summarize our approach (\cref{fig:codegen_recovery}) in three steps:

\begin{enumerate}
\item Interpret and extract contents from log records.
\item Convert the extracted contents into physical plans that can be compiled into native code (Code Generation).
\item Run the compiled machine code using the execution engine.
\end{enumerate}
We implement our approach in NoisePage~\cite{noisepage}. Noisepage depends on a recovery system that uses physical logging schemes. It also adopts query compilation optimization for its execution engine.

\section{Recovery System Architecture}
 To show how we will extend the recovery system, we first present some background information on the recovery system architecture of NoisePage. NoisePage's recovery system consists of three major components: Transaction Manager, Logging Manager, and Recovery Manager. Both the Logging Manager and the Recovery Manager interact with the Transaction Manager to achieve logging and recovery respectively. We will first give an overview of NoisePage's transaction manager,

\subsection{Transaction Manager}
The Transaction Manager is responsible for creating and maintaining database transactions. NoisePage uses a transaction to perform physical updates on the database and physical logging to disk. NoisePage uses a multi-versioned delta store Transaction Manager that ensures \texttt{SNAPSHOT-ISOLATION}. The Transaction Manager allows non-blocking reads over writes and vice versa, but does not allow write-write conflicts on a per tuple basis. A transaction's lifecycle begins when the transaction begins, and ends after the log manager has serialized its changes to disk.

Each data tuple in NoisePage is uniquely identified through a tuple slot that stores the offset of a tuple. When the DBMS inserts a tuple into a table, it creates a new tuple slot that points to a memory location within the table. Subsequent updates to the tuple does not create direct copies of the tuple, but stores delta information (redo records) about the updates instead. These redo records are essential for executing log replay correctly. The structure of a redo record is shown later in \cref{fig:log_record}.

A transaction uses a buffer (redo buffer) that stores all the delta records that will appear in its lifetime. Each redo buffer has a fixed size and is allocated from a centralized memory pool. Each time a transaction needs to change the contents of the DBMS, it will attempt to request space from the redo buffer to dump a corresponding delta record of the change. If the redo buffer runs out of space, the transaction will replace the current buffer with a new buffer from the memory pool. The redo buffer allows the logging components to process the changes before the transaction is completed.

When a transaction commits, the Transaction Manager creates a commit record that contains a timestamp that refers to the oldest active transaction from the timestamp manager. The commit record is appended to the redo buffer. The redo buffer is then carried over to the log manager, where the changes will be serialized to disk. When a transaction aborts, the transaction creates an abort record that will prevent corresponding records to be replayed during recovery. Abort records are essential; since the redo buffer is persisted to disk by the log manager once it is full, it is possible for aborted transactions to persist on disk. Aborted transactions that are persisted on disk are rare, however, since it would be difficult for a typical OLTP transaction to fill up the entire redo buffer.

@@

The Log Manager (\cref{fig:log_manager}) is responsible for performing write-ahead logging in NoisePage. All the changes are serialized to disk as logs before they are performed on the DBMS storage. Each log record is self-contained and does not require any additional metadata to be replayed. The Log Manager consists of two separate tasks that live in different threads: the Log Serializer tasks and the Log Consumer tasks.

\subsubsection{Log Serializer}
The Log Serializer task receives redo buffers from a transaction once it is full, or when the transaction decides to commit. The Log Serializer breaks down the redo buffers into raw memory segmented into fixed-size buffers. These buffers will be consumed by the Log Consumer for persistence.

The log records must be serialized in the correct order for recovery to perform correctly. Different transactions can share a same redo buffer that is filled with each of their own log records. Moreover, it is possible for transactions to appear in a non-serial order relative to their begin timestamp. However, the log records performed by an individual transaction is guaranteed to appear in the same order as they were created. This property helps ensure \texttt{SNAPSHOT-ISOLATION} property of NoisePage during log replay.
\subsubsection{Log Consumer}
Buffers supplied by the log serializer can be consumed by different Log Consumer tasks. Each Log Consumer is given a copy of the original buffer to achieve parallelism. NoisePage uses two type of log consumers: Disk Log Consumer and Replication Log Consumer.

@@

\cref{fig:recovery_manager} shows the architecture of the Recovery Manager that is responsible for log replay. NoisePage uses four different record types: redo, delete, commit, abort. Redo and delete records are not processed immediately. Instead, they are placed inside a deferred changes buffer, ordered by their transaction timestamp. An abort record removes the corresponding redo/delete record from the buffered changes. A commit record initiates a transaction to iterate over the deferred changes buffer. Once all the deferred changes are processed, the buffer is cleared and the transaction is committed. The transaction aborts if any log record is malformed or any replay fails.

@@

A physical log record consists of a header followed by a record body (\cref{fig:log_record}). The header stores metadata about the log record itself, including its record type, size and begin timestamp of the transaction that generated the log record as mentioned in Section 3.1.1. The record body contains physical information that is necessary for recovery and varies depending on the record type. Depending on the value type, the values may bee stored in-line or not in-inline. Simple types such as \texttt{INTEGER}, \texttt{FLOAT} can be stored in-line, while \texttt{VARLEN} entry may store a pointer in the values column that points to another physical location in the record.

Tuple slots in the log records will no longer be valid memory locations during recovery. Instead, the Recovery Manager creates an internal mapping from original tuple slots to a new tuple slots that will be retrieved after new values are inserted. The mapping allows the Recovery Manager to apply changes to the correct memory locations in the new memory environment.

Since only redo and delete records will effectively modify physical storage, we only need to analyze the record body of those two record types. Modifications of a query to the DBMS can be resolved down to the modification of tuple slots on the physical level.

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/Delta.png}
\caption{Redo Record Projected Row.}
\label{fig:projected_row}
\end{figure}

 While storing tuple slots is sufficient for delete records, redo records require additional data to perform log replay. These data are kept in the record projected row (\cref{fig:projected_row}), which stores a mapping from column ids to values. The DBMS retrieves individual values by computing offsets.

\subsubsection{Replay Step}
A replay step refers to the recovery manager replaying a single redo record or delete record. A NoisePage replay step has three different scenarios:
\begin{itemize}
    \item \textbf{Insert Replay}: The Recovery Manager inserts the values stored within a redo record into the given tuple slot and updates the indexes.
    \item \textbf{Delete Replay}: The Recovery Manager deletes the tuple slot specified by a delete record and updates the indexes.
    \item \textbf{Update Replay}: The Recovery Manager updates the tuple slot using the values within a redo record and updates the indexes.
\end{itemize}

While both insert and update redo records are represented using the same redo record structure, they can be differentiated by keeping track of all the tuple slots found in log records. If a tuple slot inside a redo record is unseen before, then the redo record represents an Insert operation. Otherwise, the Redo Record represents an Update operation.

\section{Recovery Code Generation}
The code generation structure in NoisePage follows from the data-centric compilation approach in \cite{neumann11}, as seen in \cref{fig:query_compilation}. A query is interpreted, optimized and converted into a physical plan tree. A plan tree consists of plan nodes and specifies how the query should be executed on the physical level of the database.

A plan translator then converts a physical plan tree/node into an operator pipeline. Each plan translator correspond to a plan node in the original plan tree. A pipeline contains an ordered collection of relational operators that are not yet materialized, ended with a pipeline breaker (e.g., sort tuples, creating a hash table). The pipeline is only materialized whenever a pipeline breaker has been encountered.

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/LLVM.png}
\caption{Lifecycle of a TPL code fragment. Execution can run in either interpret or compile mode.}
\label{fig:llvm}
\end{figure}
NoisePage then compiles each operator pipeline into a customized imperative language referred to as (TPL). The DBMS then compiles each function of a query inside the TPL code into some bytecode (TBC). The bytecode can be interpreted by a built-in virtual machine, or further compiled into an LLVM module. The execution engine can either interpret TBC, or run the compiled LLVM module (\cref{fig:llvm}).

@@

We further demonstrate the process of query compilation through the example in \cref{fig:codegen_example}. To run the query \dbSQL{DELETE FROM x WHERE col BETWEEN a AND b}, the DBMS needs to first do a sequential scan to find out the columns, then do a delete operation on the found columns. These two steps are manifested in the delete plan node that includes a sequential scan plan node as a child. The plan node then translates into an operator pipeline that specifies how each operation will be exactly executed in the DBMS. The operator pipeline then converts to TPL code. Each function inside the TPL code correspond to a piece of function inside the execution engine that will be evetually invoked once the compiled native code is run. For instance, \texttt{@tableDelete} corresponds to a C++ function \texttt{OpStorageInterfaceTableDelete} (a delete operation on a tuple slot of a table) in the execution engine.

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/ExpressionGeneration.png}
\caption{Process for converting log records into native code.}
\label{fig:expression_generation}
\end{figure}
We imagine each replay as a form of SQL statement that can be converted into native code through the code generation process discussed above (\cref{fig:expression_generation}). The code generation process for a log record can be divided into two steps: plan generation (creating the physical plan node) and plan translation (translating the plan into pipelines). Each replay (Insert Replay, Update Replay, Delete Replay) will be matched to a corresponding plan node.

\subsection{Insert Replay Conversion}

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/InsertConversion.png}
\caption{Conversion from an Insert Redo Record into an Insert Plan Node. The tuple slot provided by the redo record is discarded.}
\label{fig:insert_conversion}
\end{figure}
Insert plan nodes correspond to \dbSQL{INSERT} statements. We can represent an insert redo record as the SQL statement \dbSQL{INSERT INTO x VALUES y}. The insert plan node for this statement does not require any child plan nodes to function (\cref{fig:insert_conversion}). This makes it straightforward to convert from an insert redo record to a plan node. Index information and column ids can be retrieved from the catalog. Values stored in the redo records are in their most basic form (e.g., integer, float, double) as constants and can be converted into expressions to be accepted by the insert plan node. The operation of transferring values, however, require one-to-one copy of each value from the log record into the plan node.

\subsection{Delete Replay Conversion}
\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/DeleteConversion.png}
\caption{Conversion from a Delete Redo Record into a Delete Plan Node. Tuple slot that will be used in the pipeline is "smuggled" from the redo record through Execution Context.}
\label{fig:delete_conversion}
\end{figure}

Delete plan nodes correspond to \dbSQL{DELETE} statements and require child nodes to function. A delete plan node uses the child plan node to figure out which tuple slot to schedule for deletion. This is different for delete records. The mapping from Recovery Manager combined with the original tuple slot already points to the tuple slot to delete. In this case, there will be no need for the plan node to go through the child node and resolve the delete tuple slot. This is because, eventually, the delete plan node will attempt to create a Delete Record, which is structurally identical to the Delete Record from the Recovery Manager. We will therefore change the behavior of the operator pipeline to retrieve the delete slot supplied by the recovery manager during recovery (\cref{fig:delete_conversion}), and remove the step to resolve which tuple slot to delete. Other operations, such as index deletions, are still necessary and are kept in place.

\subsection{Update Replay Conversion}
Update plan nodes correspond to \dbSQL{Update} statements and require child nodes to function. Unfortunately, update plan nodes are not very compatible with update redo records. After an update plan is translated into an update operator pipeline, it expects set clauses returned from the child node (e.g., sequential scan). These set clauses, however, describe higher-level logical operations and cannot be constructed as easily and efficiently as creating primitive expressions for insert plan nodes. 
\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/UpdateConversion.png}
\caption{Conversion from an Update Redo Record into an Update Plan Node. Set clauses and child plan nodes inside the update plan node are discarded.}
\label{fig:update_conversion}
\end{figure}
The update plan node therefore needs to be extended to accept primitive values like an insert plan node. We also modify the operator pipeline to process the primitive values instead of set clauses during a replay step (\cref{fig:update_conversion}). Functions inside the operator pipeline will now apply to the primitive value expressions. Similar to delete record transformations, there is no need for the update pipeline to resolve the update tuple slot or to create a new update record, as these information already exist in an update redo record and the tuple slot mapping in the Recovery Manager.

\section{Recovery Execution}
The execution engine is reponsible for running the compiled query. After we have finished the setup for compiling different replays into native code, we will explain how they will affect the execution engine.

\subsection{Storage Interface}
The compiled query invokes different operation codes (e.g., \texttt{OpStorageInterfaceTableDelete} in \cref{fig:codegen_example}) to complete the corresponding operator in the TPL code. Executing the operation codes requires access to the storage layer of the DBMS. In NoisePage, the storage layer is separate from the execution engine. Therefore, it uses a Storage Interface to interact with contents in the storage layer. The Storage Interface stores objects that allow manipulation over a certain DBMS table. These objects include an Execution Context, a redo record that will be used to persist changes done to the table, projected rows associated with the redo record and index objects associated with table. A Storage Interface is reconstructed every time an execution of a query begins, and is destructed after the query has been executed. We will need to modify functions that rely on the Storage Interface and access members that reside within the Storage Interface.

\subsection{Execution Changes}
\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/RecoveryExecution.png}
\caption{Information from log record can be retrieved from the Execution Context.}
\label{fig:recovery_execution}
\end{figure}

We explained in Section 3.2.2 and 3.2.3 that for our purposes, we expect deletes/updates to retrieve relevant tuple slots from the Recovery Manager. However, the built-in pipeline assumes that the tuple slots are retrieved later during execution. We can circumvent this rule by modifying operation code functions and the execution context. The operation code functions provide access to the Storage Interface that is required to access the execution context.

An Execution Context encapsulates information that is supplied by upper layers of the DBMS. These information include access to the database catalog, the transaction running the query, etc. We can also add tuple slot information into the Execution Context. Since the Execution Context can be accessed throughout a Storage Interface, we can modify the behavior of operation codes to override with the recovery tuple slots, as shown in \cref{fig:recovery_execution}.

>>>>

\hypersetup{colorlinks,linkcolor={black},citecolor={blue},urlcolor={black}}

@@

\begin{abstract}
Code generation is an optimization technique for improving query execution throughput by compiling query plans into native code. This technique, however, leads to design challenges for the recovery system of a database management system. The log replay process will be disjoint from the built-in execution engine that has adapted to operate efficiently on compiled code. This leads to the implementation of a separate execution engine to deal with the execution of log records, which can be a huge waste of engineering efforts. To resolve this design conflict between code generation and database recovery, we present a unified approach to support both query execution and log replay in a code-generation-based DBMS. We ask the recovery system to convert log records into compiled code that will be easily accepted by the execution engine. Our results show that while our approach incurs a higher performance overhead compared to using a separate execution engine, it requires much less engineering effort and is superior in index updates.
\end{abstract}

@@

Moreover, the DBMS recovery system is isolated from the query compilation execution engine (i.e., a component of a DBMS that is responsible for execution). Most of the query compilation DBMSs we have surveyed~\cite{noisepage, neumann11, paroski16, craig14} have no conduits between their recovery systems and execution pipelines. The recovery system implements its own functionalities to update the database, but some of these functionalities already exist in the execution engine. For instance, we mentioned that in-memory DBMSs rebuild the indexes during log replay. When a redo process install changes on a table, it also retrieves the indexes on that table and update them (index update). However, the execution engine has already implemented this index update functionality. As a result, the DBMS contains two implementations that provide identical functionalities. The number of those functionalities in the recovery system will eventually amount to the scale of a separate execution engine that is built specifically for log replay.

@@

@@

There are two solutions to this problem. One solution is to re-implement functions to install physical changes on data tables within the recovery system. However, this approach completely isolates the recovery system from the execution engine. This means that if the recovery system wishes to use a functionality within the execution engine, it has to rewrite that same functionality within the recovery system (\cref{fig:recovery_index_update}).

@@

If the DBMS can convert log records into physical plans, the recovery system can rely on code generation for recovery operation. This code generation recovery approach brings around several benefits. Firstly, it significantly reduces engineering overhead. Compiling contents of the log records into machine code allows the recovery system to seamlessly integrate with the built-in execution engine. We have explained in Section 2.1 that the recovery system needs to perform extra operations (e.g. \cref{fig:recovery_index_update}) besides updates to the data tables. These functionalities exist within the execution engine. If we allow the recovery system to reuse existing functionalities by communicating with the execution engine through native code, then the recovery system no longer needs to implement its own functionalities. Secondly, utilizing the code generation execution engine provides more efficient functionalities. A function implemented by the recovery system will not be as efficient as one that exists in the execution engine.

These potential benefits motivate us to design a code generation recovery system based on conversion from log records to physical plans. In the next chapter, we will present the implementation of our approach .

@@

We now describe how to extend the recovery system and query compilation pipeline to support log replay with code generation. Our implementation follows along the third option shown in \cref{fig:motivation}. We first enable the recovery system to convert log records into compiled native code. To achieve this, the DBMS needs to interpret the log records into physical plans and prepare them for compilation. We then need to make minor modifications to the execution engine, so it will be able to correctly execute the native code generated by the recovery system. We further summarize our approach (\cref{fig:codegen_recovery}) in two steps:

\begin{enumerate}
\item Convert the log records into physical plans that can be compiled into native code (Recovery Code Generation).
\item Run the compiled machine code using the execution engine (Recovery Execution) for log replay.
\end{enumerate}
We implement our approach in NoisePage~\cite{noisepage}. Noisepage depends on a recovery system that uses physical logging schemes. It also adopts query compilation optimization for its execution engine.

\section{Recovery System Architecture}
 To show how we will extend the recovery system, we first present some background information on the recovery system architecture of NoisePage. NoisePage's recovery system consists of three major components: Transaction Manager, Logging Manager, and Recovery Manager. The Logging Manager interacts with the Transaction Manager to achieve physical logging, while the Recovery Manager interacts with the Transaction Manager to provide database recovery.

\subsection{Transaction Manager}
The Transaction Manager is responsible for creating and maintaining database transactions. NoisePage uses a transaction to perform physical updates on the database and physical logging to disk. NoisePage uses a multi-versioned delta store~\cite{reed78} Transaction Manager that ensures \texttt{SNAPSHOT-ISOLATION}~\cite{berenson95}. The Transaction Manager allows non-blocking reads over writes and vice versa, but does not allow write-write conflicts on a per tuple basis. A transaction's lifecycle starts when the transaction begins, and ends after the log manager has serialized its changes to disk.

Each data tuple in NoisePage is uniquely identified through a tuple slot that stores the offset of a tuple in a data block. When the DBMS inserts a tuple into a table, it creates a new tuple slot that points to a memory location within the table's data block. Subsequent updates to the tuple does not create direct copies of the tuple, but stores delta information (redo records) about the updates instead. These redo records are essential for executing log replay correctly. The structure of a redo record is shown later in \cref{fig:log_record}.

A transaction uses a buffer (redo buffer) that stores all the delta records that will be generated in its lifetime. Each redo buffer has a fixed size and is allocated from a centralized memory pool. Each time a transaction needs to change the contents of the DBMS, it will attempt to request space from the redo buffer to store a corresponding delta record of the change. If the redo buffer runs out of space, the transaction will replace the current buffer with a new buffer from the memory pool. The redo buffer allows the logging components to process the changes before the transaction ends.

When a transaction commits, the Transaction Manager creates a commit record that contains a timestamp that refers to the oldest active transaction from the timestamp manager. The commit record is appended to the redo buffer. The redo buffer is then carried over to the log manager, where the changes will be serialized to disk. When a transaction aborts, the transaction creates an abort record that prevents corresponding records to be replayed during recovery. Abort records are essential to recovery; since the redo buffer is persisted to disk by the log manager once it is full, it is possible for aborted transactions to persist on disk. During recovery, the DBMS needs to remove aborted transactions. Aborted transactions that are persisted on disk are rare~\cite{tpc15}, however, since it would be difficult for an OLTP transaction to fill up the entire redo buffer.

@@

The Log Manager (\cref{fig:log_manager}) is responsible for performing write-ahead logging~\cite{aries} in NoisePage. Each log record is self-contained and does not require any additional metadata to be replayed. The Log Manager consists of two separate tasks that live in different threads: the Log Serializer tasks and the Log Consumer tasks.

\subsubsection{Log Serializer}
The Log Serializer task receives redo buffers from a transaction once it is full, or when the transaction decides to commit. The Log Serializer breaks down the redo buffers into raw memory segmented into fixed-size buffers. The Log Consumer consumes these buffers for persistence.

The DBMS must serialize log records in the correct order for recovery to perform correctly. Different transactions can share a same redo buffer that is filled with each of their own log records. Moreover, it is possible for transactions to appear in a non-serial order relative to their begin timestamp. However, the log records performed by an individual transaction is guaranteed to appear in the same order as they were created. This property helps ensure \texttt{SNAPSHOT-ISOLATION} property of NoisePage during log replay.
\subsubsection{Log Consumer}
Different Log Consumers are responsible for consuming buffers supplied by the Log Serializer. The Log Serializer provides each Log Consumer a copy of the original buffer to achieve parallelism. NoisePage uses two type of log consumers: Disk Log Consumer and Replication Log Consumer.

@@

\cref{fig:recovery_manager} shows the architecture of the Recovery Manager that is responsible for log replay. The Recovery Manager loads the log records into memory and use them to performs log replay. NoisePage uses four different record types: redo, delete, commit, abort. The Recovery Manager do not process redo and delete records are immediately. Instead, the Recovery Manager place redo and delete records inside a deferred changes buffer, ordered by their transaction timestamp. An abort record removes a corresponding redo/delete record from the buffer using the begin transaction timestamp (\cref{fig:log_record}). A commit record initiates a transaction to iterate over the deferred changes buffer. Once all the Recovery Manager processed all the deferred changes, it clears the buffer and commits the transaction. The transaction aborts if any log record is malformed or any replay fails.

@@

A physical log record consists of a header, followed by a record body (\cref{fig:log_record}). The header stores metadata about the log record itself. This includes a log record's record type, size, and begin timestamp of the transaction (See Section 3.1.1). The record body contains physical information that is necessary for recovery and varies depending on the record type. Depending on the value type, the values may bee stored in-line or not in-inline. Simple types such as \texttt{INTEGER}, \texttt{FLOAT} can be stored in-line, while \texttt{VARLEN} entry may store a pointer in the values column that points to another physical location in the record.

Tuple slots in the log records will no longer be valid memory locations during recovery. Instead, the Recovery Manager creates an internal mapping from original tuple slots to new tuple slots that will be retrieved after new values are inserted. The tuple slot mapping allows the Recovery Manager to apply changes to the correct memory locations in the new memory environment.

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/Delta.png}
\caption{Redo Record Projected Row.}
\label{fig:projected_row}
\end{figure}
\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/PRExample.png}
\caption{Projected Row that represents [15, 721, \dbSQL{NULL}].}
\label{fig:projected_row_example}
\end{figure}

 While storing tuple slots is sufficient for delete records, redo records require additional data to perform log replay. For instance, to perform an insert, the redo record must contain the values to insert. The Log Manager stores these data in a log record's projected row (\cref{fig:projected_row}). A projected row represents a row in a database table. The DBMS can access an individual value within the projected row by identifying a column id and reading the value using that column's value offset (\cref{fig:projected_row_example}).

\subsubsection{Replay Step}
The Recovery Manager processes each log record differently depending on its record type. Commit and abort records do not change the physical storage of the the database. Therefore, to simplify, we only need to consider the effects of redo and delete records.

We refer to a replay step as the Recovery manager replaying a single redo record or delete record. A NoisePage replay step has three different scenarios:
\begin{itemize}
    \item \textbf{Insert Replay}: The Recovery Manager inserts the values stored within a redo record into the given tuple slot and updates the indexes.
    \item \textbf{Delete Replay}: The Recovery Manager deletes the tuple slot specified by a delete record and updates the indexes.
    \item \textbf{Update Replay}: The Recovery Manager updates the tuple slot using the values within a redo record and updates the indexes.
\end{itemize}

While both insert and update redo records follow the same redo record structure, the Recovery Manager differentiates between them by keeping track of all the tuple slots found in log records. If the Recovery Manager has never seen a tuple slot inside a redo record before, then the redo record represents an Insert operation. Otherwise, the Redo Record represents an Update operation.

\section{Recovery Code Generation}
In this section, we introduce how we integrate the NoisePage's recovery system with its code generation pipeline. We first provide an overview over NoisePage's code generation pipeline. We then show how we convert redo and delete records into physical plans that can be accepted by the code generation pipeline for compilation.

\subsection{Code Generation Pipeline}
The DBMS interprets a query, parses it into an AST and converts it into a physical plan tree. A plan tree consists of plan nodes and specifies how the query should be executed on the physical level of the database.

Code generation is responsible for compiling physical plans into machine code. The code generation pipeline in NoisePage follows from the data-centric compilation approach\cite{neumann11}, as shown in \cref{fig:query_compilation}. The DBMS feeds each plan node into a matching plan translator. Each plan translator correspond to a plan node in the original plan tree. For instance, an insert plan node has a matching insert plan translator; a delete plan node has a matching delete plan translator. A plan translator converts a physical plan tree/node into an operator pipeline. The operator pipeline contains an ordered collection of relational operators (e.g. insert, get table row) that are not yet materialized, ended with a pipeline breaker (e.g., sort tuples, creating a hash table). The execution engine only materializes the pipeline results when a pipeline breaker has been encountered.

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/LLVM.png}
\caption{Lifecycle of a TPL code fragment. Execution can run in either interpret or compile mode.}
\label{fig:llvm}
\end{figure}
NoisePage then compiles each operator pipeline into a customized imperative language (TPL). The DBMS then compiles each function of a query inside TPL into some bytecode (TBC). The bytecode can be interpreted by a built-in virtual machine, or further compiled into an LLVM module. The execution engine can either interpret TBC, or run the compiled LLVM module (\cref{fig:llvm}).

@@

We further demonstrate the process of query compilation through the example in \cref{fig:codegen_example}. To run the query \dbSQL{DELETE FROM x WHERE col BETWEEN a AND b}, the DBMS needs to first do a sequential scan to find out the columns, then do a delete operation on the found columns. The DBMS represents these two operations in a delete plan node that includes a sequential scan plan node as a child. The DBMS then translates the plan node into an operator pipeline that specifies how each operation will be exactly executed in the DBMS. After this, the DBMS compiles the operator pipeline into machine code (TPL, TBC, LLVM). Each function inside the machine code matches to an operation code function inside the execution engine. When the execution engine runs the compiled code, it invokes its corresponding operation code functions. For instance, \texttt{@tableDelete} corresponds to a C++ function \texttt{OpStorageInterfaceTableDelete} (a delete operation on a tuple slot of a table) in the execution engine.

\subsection{Recovery Integration}
\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/ExpressionGeneration.png}
\caption{Process for converting log records into native code.}
\label{fig:expression_generation}
\end{figure}
We can consider each type of replay as a form of SQL statement that the DBMS can convert into native code through the code generation process discussed above (\cref{fig:expression_generation}). We further divide the code generation process for a log record into three steps: (1) plan generation (creating the physical plan node), (2) plan translation (translating the plan into operator pipelines), and (3) compilation.

In plan generation, the Recovery Manager converts each replay process (insert/update/delete replay) into a corresponding plan node. In plan translation, the Recovery Manager uses matching plan translators to convert physical plans into operator pipelines. Once the Recovery Manager has the operator pipeline, it can then use the compiler to generate machine code for execution.

\subsubsection{Insert Replay Conversion}
\label{insert_replay_conversion}

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/InsertConversion.png}
\caption{Conversion from an Insert Redo Record into an Insert Plan Node. The tuple slot provided by the redo record is discarded.}
\label{fig:insert_conversion}
\end{figure}
Insert plan nodes correspond to \dbSQL{INSERT} statements. We can represent an insert redo record as the SQL statement \dbSQL{INSERT INTO x VALUES y}.

\cref{fig:insert_conversion} shows the conversion from an insert redo record into an insert plan node. The Recovery Manager can retrieve metadata (e.g. table id, column ids) from the redo record and the database catalog. The redo record's projected row contains values for insert (\cref{fig:projected_row}). Those values are in primitive forms (e.g., integer, float, double) as constants. The Recovery Manager then feeds those constants into the insert plan node. This conversion requires no changes to the plan nodes, nor to the operator pipelines.

\subsubsection{Delete Replay Conversion}
\label{delete_replay_conversion}

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/DeleteConversion.png}
\caption{Conversion from a Delete Redo Record into a Delete Plan Node. The operator pipeline uses the tuple slot from the delete record.}
\label{fig:delete_conversion}
\end{figure}

Delete plan nodes correspond to \dbSQL{DELETE} statements. \dbSQL{DELETE} statements require specifications on where the DBMS should perform delete operations (e.g., \dbSQL{WHERE} clause in \cref{fig:codegen_example}). Therefore, a delete plan node requires a child plan node to function. It uses its child plan node to figure out which tuple slot to schedule for deletion.

However, a delete record's tuple slot, combined with the tuple mapping from Recovery Manager, already points to the tuple slot to delete. In this case, there will be no need for the plan node to go through the child node and find out the delete tuple slot. Moreover, delete records do not contain value information that is required for index update (\cref{fig:log_record}). Therefore, we need to provide the operator pipeline with a projected row that contains values that correspond to the row of the tuple slot from the Recovery Manager.

We change the behavior of the operator pipeline as follows (\cref{fig:delete_conversion}): (1) remove the step to look for a new delete tuple slot, (2) retrieve the delete slot supplied by the Recovery Manager during execution, and (3) use the projected row provided by the Recovery Manager for index updates.

\subsubsection{Update Replay Conversion}
\label{update_replay_conversion}

Update plan nodes correspond to \dbSQL{UPDATE} statements. Similar to \dbSQL{DELETE} statements, update plan nodes require child plan nodes. An update operator pipeline expects \dbSQL{SET} clauses returned from the child node (e.g., sequential scan). These \dbSQL{SET} clauses, however, describe higher-level logical operations (i.e., update a column with a certain value). On the other hand, redo records contain low-level physical information (i.e., update a tuple slot with a certain value). Therefore, the Recovery Manager cannot create \dbSQL{SET} clauses efficiently from redo records.

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/UpdateConversion.png}
\caption{Conversion from an Update Redo Record into an Update Plan Node. \dbSQL{SET} clauses and child plan nodes inside the update plan node are discarded.}
\label{fig:update_conversion}
\end{figure}

We therefore perform the following changes (\cref{fig:update_conversion}): (1) extend the update plan node to accept primitive values like an insert plan node does, (2) modify the operator pipeline to process the primitive values instead of \dbSQL{SET} clauses during a replay step (\cref{fig:update_conversion}), (3) remove the step for update operator pipeline to find out the update tuple slot, and (4) retrieve the update tuple slot from the recovery manager.

\section{Recovery Execution}
We explained in \cref{delete_replay_conversion} that we modified deletes/update operator pipelines to retrieve certain objects (e.g., tuple slots, projected row) from the Recovery Manager. To achieve this, we need to modify the execution engine. The execution engine is responsible for running the compiled query. Inside the execution engine, the compiled query invokes different operation codes (e.g., \texttt{OpStorageInterfaceTableDelete} in \cref{fig:codegen_example}) that correspond to operators in the operator pipeline. Since we modified the operator pipelines earlier, we now need to change operation codes to read in objects from the Recovery Manager.

\subsection{Execution Changes}
\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/RecoveryExecution.png}
\caption{Operation code functions can access log records through an Execution Context.}
\label{fig:recovery_execution}
\end{figure}

We first use the Recovery Manager to set the required tuple slots and projected row into a log record. Before the execution engine runs the query, the Recovery Manager needs to setup a preparation context for execution. In NoisePage, an execution context encapsulates information that is supplied by upper layers of the DBMS (e.g. access to the database catalog, the transaction running the query to execute, database memory pool). We pass the log record from the Recovery Manager into the execution context.
